Untwisting the Tetrahelix (v0.12)

### 2017-03-24 doc/TetrahelixUntwist.tex
updating photos
latest improvements

\title{Untwisting the Tetrahelix (v0.10)}
\author{Robert L. Read \texttt{read.robert@gmail.com} \and
  Robert Gatliff \texttt{robert@toubat.org}
}

@@

\begin{abstract}
  The Boerdijk--Coxeter helix (BC helix, or tetrahelix) is a
  face-to-face stack of regular tetrahedra.  Considering the edges of
  these tetrahedra as structural members, the resulting structure is attractive and
  inherently rigid, and therefore interesting to architects,
  mechanical engineers, and robotocists.  A formula that matches the
  visually apparent helices forming the outer rails of the BC helix.
  This formula is generalized to a formula convenient to designers.
  Formulae for 
  computing the
  parameters that give an edge-length minimax-optimal tetrahelices
  are given, defining a continuum of tetrahelices of varying curvature.
  The endpoints of the optimality of this continuum are the BC helix and
  a structure of zero curvature, the \emph{equitetrabeam}.
  Numerically finding the rail angle from the equation for
  pitch allows optimal tetrahelices of any pitch to be designed. 
  An interactive tool for such design and experimentation is provided: \url{https://pubinv.github.io/tetrahelix/}.
  A formula for the inradius of optimal tetrahelices is given.
  Utility and use for static and variable geometry
  truss/space frame design and robotics based on the are discussed.
\end{abstract}

@@

\begin{figure}[H]
  \label{fig:closeup}
  \centering
     \includegraphics[width=0.95\textwidth]{figures/BCHelixCloseUp.png}
     \caption{BC Helix Close-up}
\end{figure}

Imagining Figure \ref{fig:closeup} as a static mechanical structure,
we observe that it is useful to the mechanical engineer or
robotocist because the structure remains an inherently rigid,
omni-triangulated space frame, which is
somewhat mechanically strong.
Imagine further in Figure \ref{fig:closeup}, that each static edge was replaced with an
actuator that could dynamically become shorter or longer in response to electronic control,
and the they vertices were a joint that supported sufficient angular displacement
for this to be possible. Such a machine is a Tetrobot\cite{TetrobotBook}.

@@

We would like to design nearly regular tetrahelices with a formula that
gives the vertices in space. Eventually we would like to design nearly regular
tetrahelices by choosing the lengths of a small set of members.
If you are designing a
space frame, this is a static design choice, in a robot, it is a
dynamic choice that can be used to twist\footnote{The formal definition of twistieness, or \emph{torsion},
  is not useful or used in the paper. The formal \emph{curvature} and \emph{torsion} of the helices defined here
may be easily computed from the formualae if desired.} the robot and/or exert linear or
angular force on the environment.

@@

$\theta$ is approximately $131.81$ degrees.
The angle $\theta$ is the rotation of a \emph{each} tetrahedra, not the tetraheda along a rail.  In Figure \ref{fig:closeup},
each tetrahedra has either a yellow, blue, or red outer edge or rail.
That is, a blue-rail tetrahedron is rotated slightly more than a $1/3$ of a revolution to match the face of the yellow tetrahedra.

R.W. Gray's site\cite{graytetrahelix}, repeating a formula by Coxeter\cite{coxeter1985simplicial} in more accessible form, gives for a counter-clockwise BC Helix:
\begin{equation}
  \label{graycoxeter}
V(n) =
\left [
  \begin{tabular}{c}
   $ r_{bc} \cos{n \theta} $\\
   $ r_{bc} \sin{n \theta} $\\
   $ n h_{bc}  $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $ r_{bc} = \frac{3\sqrt{3}}{10} $\\
 $ h_{bc} = 1/\sqrt{10} $ \\
 $ \theta = \arccos(-2/3) $ \\
  \end{tabular}      
\end{equation}
where $n$ represents each integer numbered node in succession on every colored rail.

The apparent rotation of tetrhedra of the same outer-edge color ($V_3$ relative from $V_0$ for example,
in \eqref{graycoxeter}, is $3 \theta - 2\pi$).

@@

This Gray-Coxeter formulation $V(n)$ from \eqref{graycoxeter} is valuable, but obscures the fact that we have three
apparent rail helices.
In order to rewrite this expression with an explicit rotation of $2\pi/3$, we expand 
the expressions $\theta$ and $\rho_{bc}$ in \eqref{eq:colored} and seek to isolate the term $c2\pi/3 $.
\begin{align*}
 c \theta  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (c/3)  (3 \theta)  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
 (c/ 3) ((3 \theta - 2 \pi)  + 2 \pi) &= \text{\{definition of $\rho_{bc}$\}...} \\  
  (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{algebra...} \\  
c  ( \rho_{bc} +  2 \pi) /3  \\
\end{align*}
We use this to a create
a new formulation: $H_{BCsymmetric}(n,c) = H_{BCcolored}(n,c)$.
%% TODO: I seem to have lost the thread of my thought. I don't seem to be introducing d at that right time.

@@

%% Actually, I am not sure if this is a clearer formulation...it might be, and need to put it in general!!!
\begin{equation}
H_{BCsymmmetric}(n,c) =
\left [
  \begin{tabular}{c}
    $ r  \cos{\rho_{bc} n + c (\rho_{bc} +  2 \pi) /3} $\\
   $ r  \sin{\rho_{bc} n + c (\rho_{bc} +  2 \pi) /3} $\\
   $ (n+c/3)  h_{bc} $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
%%  $\kappa = n + c/3$ \\
    $\rho_{bc} = (3 \theta - 2 \pi)$ \\
    $ \theta = \arccos(-2/3) $ \\
    $ h_{bc} = 1/\sqrt{10} $ \\    
  \end{tabular}      
\end{equation}

We can see the each rail tetrahedron moves us $\rho_{bc}$ radians around a circle. Since:
\[ \frac{2 \pi}{\rho_{bc}} \approx 10.16
\]
we can see that there are approximately $10.16$ red, blue or yellow tetrahedra on one rail in a single revolution.

@@

We use the term \emph{tetrahelix} to mean any structure made of
vertices and edges which is isomorphic to the BC helix, in which the
vertices lie on three helices. We further demand that all edge lengths
be finite, as we are only interested in physically constructable tetrahelices.
By isomporphic we mean there is a one-to-one mapping between both
vertices and edges in the two tetrahelices.
One could consider various definitions of optimality for a
tetrahelix, but the must useful to an enginner is to minimize the
maximum difference between any two edges.

@@

Since the axes are parallel, we may define the \emph{inradius} of a
tetrahelix to be the radius of the largest
cylinder parallel to this axis contained within the circumscribing cylinder
which is penetrated by no edge.
Define a \emph{minimax edge-length optimal tetrahelix} or just an
\emph{optimal tetrahelix} to be a tetrahelix for which there exists
no other tetrahelix of the same inradius and pitch with a lower maximum
difference between its edge lengths.

@@

Since all three rails have the same rail length, no matter how we
move the rails in the $xy$ plane if we shorten the $xy$ distance between
vertices we shorten the total distance.
Our tool for thinking about this is to project out the $z$ dimension to form
a two-dimensional figure of nodes and non-rail edges (see Figure \ref{untouchingrailfig}.
Consider the projection along the $z$ axis of all vertices and non-rail edges into the $xy$ plane, which will be
a figure of dots and connecting segments in the $xy$ plane. The convex
hull for any one helix projection will be a circle (if its pitch is
irrational) or a polygon if rational, or a point if the helix has
zero curvature. Each of these figures by definitions lies outside the
circle of inradius $i$ in the $xy$ plane.

@@

$H_{general}$ forces the user to select a $d_{\rho}$
which has a sensible radius.

@@

If you look down the axis of an optimal tetrahelix, it happens that only one of the one-hop edges
come closest to the axis. In other words, they define the radius of the incircle of the
projection, or the radius of a cylinder that would just fit inside the tetrahelix.
This is useful to know if the tetrahelix is used as a structure to house a pipe or
a ladder that must bear a human or to armor a conduit. The inradius $i_\rho$ of
an optimal tetrahelix is a remarkably simple function of the radius $r$ and the rail angle $\rho$:
\begin{equation}
  \label{eq:inradius}
  i_{\rho} = r \sin{\frac{\pi - \rho}{6}}
\end{equation}
Which can be seen from the trigonometry of a diagram of the projected one-hop edges
connecting four sequentially numbered vertices:

@@

In Figure \ref{fig:spudboy}, the furthest tetrahelix is the optimal equitetrabeam.
Figure \ref{fig:unfolded} shows a representation of an unfolded equitetrabeam.

@@

As suggested by Buckminster Fuller, the most convenient geometries to
consider are those that have regular member lengths, in order to
facilitate the inexpensive manufacture and construction of the robot.
In a plane, the octet truss is such a geometry, but in a line, the
Boerdijk--Coxeter helix is a regular structure.

@@

\begin{proof}
  Proof by our computer program that does this by forming a linear interpolation of links.
  See video on website\cite{readtetrahelix}.
\end{proof}

\begin{figure}[H] %float with two figures
  \centering
     \includegraphics[width=0.4\textwidth]{figures/Tetrahelix1.png}
     \caption{2/3rd Twisted Tetrahelix}
     \includegraphics[width=0.4\textwidth]{figures/Tetrahelix2.png}
     \caption{1/3rd Twisted, 2/3rd Untwisted Tetrahelix}
     \includegraphics[width=0.4\textwidth]{figures/Tetrahelix3.png}
     \caption{The Equitetrabream: Fully Untwisted Tetrahelix}
\end{figure}

>>>>

\title{Untwisting the Tetrahelix (v0.12)}
\author{Robert L. Read \texttt{read.robert@gmail.com} \and
  Robert Gatliff \texttt{robert@toubat.org}
}

@@

\begin{abstract}
  The Boerdijk--Coxeter helix (BC helix, or tetrahelix) is a
  face-to-face stack of regular tetrahedra.  Considering the edges of
  these tetrahedra as structural members, the resulting structure is attractive and
  inherently rigid, and therefore interesting to architects,
  mechanical engineers, and robotocists.  A formula is developed that matches the
  visually apparent helices forming the outer rails of the BC helix.
  This formula is generalized to a formula convenient to designers.
  Formulae for 
  computing the
  parameters that give edge-length minimax-optimal tetrahelices
  are given, defining a continuum of tetrahelices of varying curvature.
  The endpoints of the optimality of this continuum are the BC helix and
  a structure of zero curvature, the \emph{equitetrabeam}.
  Numerically finding the rail angle from the equation for
  pitch allows optimal tetrahelices of any pitch to be designed. 
  An interactive tool for such design and experimentation is provided: \url{https://pubinv.github.io/tetrahelix/}.
  A formula for the inradius of optimal tetrahelices is given.
  Utility for static and variable geometry
  truss/space frame design and robotics is discussed.
\end{abstract}

@@

\begin{figure}[H]
  \label{fig:closeup}
  \centering
     \includegraphics[width=0.95\textwidth]{figures/BCHelixCloseUp.png}
     \caption{BC Helix Close-up (partly along axis)}
  \centering
     \includegraphics[width=0.95\textwidth]{figures/VerticalCloseUp.png}
     \caption{BC Helix Close-up (orthogonal)}
\end{figure}

Imagining Figure \ref{fig:closeup} as a static mechanical structure,
we observe that it is useful to the mechanical engineer or
robotocist because the structure remains an inherently rigid,
omni-triangulated space frame, which is
somewhat mechanically strong.
Imagine further in Figure \ref{fig:closeup}, that each static edge was replaced with an
actuator that could dynamically become shorter or longer in response to electronic control,
and the vertices were a joint that supported sufficient angular displacement
for this to be possible. Such a machine is a Tetrobot\cite{TetrobotBook}.

@@

We would like to design nearly regular tetrahelices with a formula that
gives the vertices in space. Eventually we would like to design nearly regular
tetrahelices by choosing the lengths of a small set of members.
If you are designing a
space frame, this is a static design choice, in a robot, it is a
dynamic choice that can be used to twist\footnote{The formal definition of twistiness, or \emph{torsion},
  is not useful or used in the paper. The formal \emph{curvature} and \emph{torsion} of the helices defined here
may be easily computed from the formualae if desired.} the robot and/or exert linear or
angular force on the environment.

@@

$\theta$ is approximately $131.81$ degrees.
The angle $\theta$ is the rotation of \emph{each} tetrahedra, not the tetraheda along a rail.  In Figure \ref{fig:closeup},
each tetrahedra has either a yellow, blue, or red outer edge or rail.
That is, a blue-rail tetrahedron is rotated slightly more than a $1/3$ of a revolution to match the face of the yellow tetrahedra.

R.W. Gray's site\cite{graytetrahelix}, repeating a formula by Coxeter\cite{coxeter1985simplicial} in more accessible form, gives the Cartesian coordinates $\begin{bmatrix}
           x \\
           y \\
           z
         \end{bmatrix}$
for a counter-clockwise BC Helix:
\begin{equation}
  \label{graycoxeter}
V(n) =
\left [
  \begin{tabular}{c}
   $ r_{bc} \cos{n \theta} $\\
   $ r_{bc} \sin{n \theta} $\\
   $ n h_{bc}  $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
 $ r_{bc} = \frac{3\sqrt{3}}{10} $\\
 $ h_{bc} = 1/\sqrt{10} $ \\
 $ \theta = \arccos(-2/3) $ \\
  \end{tabular}      
\end{equation}
where $n$ represents each integer numbered node in succession on every colored rail.

The apparent rotation of a vertex an outer-edge, $V(n)$ relative from $V(n+3)$ for any integer $n$
in \eqref{graycoxeter}, is $3 \theta - 2\pi$).

@@

The $H_{BCcolored}(n,c)$ formulation can be further clarified by rewriting directly in terms of the rail angle $\rho{bc}$ rather than $\theta$.
Intuitively we see an expression where $c/3$ is multiplied by a $1/3$ rotation plus the rail angle $\rho$.
We expand 
the expressions $\theta$ and $\rho_{bc}$ in \eqref{eq:colored} and seek to isolate the term $c2\pi/3 $.
\begin{align*}
 c \theta  &=   \text{\{we aim for 3 in denominator, so we split...\}} \\
    (c/3)  (3 \theta)  &=   \text{\{we want $2\pi$ in numerator, so add canceling terms...\}} \\
 (c/ 3) ((3 \theta - 2 \pi)  + 2 \pi) &= \text{\{definition of $\rho_{bc}$\}...} \\  
  (c / 3) \rho_{bc}  + c 2 \pi /3 &=  \text{algebra...} \\  
c  ( \rho_{bc} +  2 \pi) /3  \\
\end{align*}

This allows us to redefine:

%% Actually, I am not sure if this is a clearer formulation...it might be, and need to put it in general!!!
\begin{equation}
H_{BCcolored}(n,c) =
\left [
  \begin{tabular}{c}
    $ r  \cos{\rho_{bc} n + c (\rho_{bc} +  2 \pi) /3} $\\
   $ r  \sin{\rho_{bc} n + c (\rho_{bc} +  2 \pi) /3} $\\
   $ (n+c/3)  h_{bc} $
  \end{tabular}
  \right ],
\text{where:}
  \begin{tabular}{c}
%%  $\kappa = n + c/3$ \\
    $\rho_{bc} = (3 \theta - 2 \pi)$ \\
    $ \theta = \arccos(-2/3) $ \\
    $ h_{bc} = 1/\sqrt{10} $ \\    
  \end{tabular}      
\end{equation}

@@

From this formulation it is easy to see that moving one vertex on a rail moves us $\rho_{bc}$ radians around a circle. Since:
\[ \frac{2 \pi}{\rho_{bc}} \approx 10.16
\]
we can see that there are approximately $10.16$ red, blue or yellow tetrahedra on one rail in a single revolution.

@@

We use the term \emph{tetrahelix} to mean any structure made of
vertices and edges which is isomorphic to the BC helix, in which the
vertices lie on three helices. We further demand that all edge lengths
be finite, as we are only interested in physically constructable tetrahelices.
By isomporphic we mean there is a one-to-one mapping between both
vertices and edges in the two tetrahelices.
One could consider various definitions of optimality for a
tetrahelix, but the most useful to us as robotocists extending the Tetrobot concept is to minimize the
maximum difference between any two edges.

@@

Since the axes are parallel, we may define the \emph{inradius}, represented by the letter $i$, of a
tetrahelix to be the radius of the largest
cylinder parallel to this axis contained within the circumscribing cylinder
which is penetrated by no edge.
Define a \emph{minimax edge-length optimal tetrahelix} or just an
\emph{optimal tetrahelix} to be a tetrahelix for which there exists
no other tetrahelix of the same inradius and pitch with a lower maximum
difference between its edge lengths.

@@

Since all three rails have the same rail length, no matter how we
move the rails in the $xy$ plane if we shorten the $xy$ distance between
vertices we shorten the total distance.
Our tool for thinking about this is to project out the $z$ dimension to form
a two-dimensional figure of nodes and non-rail edges (see Figure \ref{untouchingrailfig}.)
Consider the projection along the $z$ axis of all vertices and non-rail edges into the $xy$ plane, which will be
a figure of dots and connecting segments in the $xy$ plane. The convex
hull for any one helix projection will be a circle (if its pitch is
irrational) or a polygon if rational, or a point if the helix has
zero curvature. Each of these figures by definitions lies outside the
circle of inradius $i$ in the $xy$ plane.

@@

$H_{general}$ forces the user to select three values: $\rho$, $r_{\rho}$, and $d_{\rho}$ satisfying \eqref{railangle}.

@@

If you look down the axis of an optimal tetrahelix, it happens that only one of the one-hop edges
comes closest to the axis. In other words, they define the radius of the incircle of the
projection, or the radius of a cylinder that would just fit inside the tetrahelix.
This is useful to know if the tetrahelix is used as a structure to house a pipe or
a ladder that must bear a human or to armor a conduit. The inradius $i_\rho$ of
an optimal tetrahelix is a remarkably simple function of the radius $r$ and the rail angle $\rho$:
\begin{equation}
  \label{eq:inradius}
  i_{\rho} = r \sin{\frac{\pi - \rho}{6}}
\end{equation}
Which can be seen from the trigonometry of a diagram of the projected one-hop edges
connecting four sequentially numbered vertices:

@@

In Figure \ref{fig:spudboy}, the furthest tetrahelix is the optimal equitetrabeam.
Figure \ref{fig:unfolded} shows a representation of an unfolded equitetrabeam.
Note that folding the outer edges of the paper upward or downward effectively
chooses the chirality of the equitetrabeam.

@@

As suggested by Buckminster Fuller, the most convenient geometries to
consider are those that have regular member lengths, in order to
facilitate the inexpensive manufacture and construction of the robot.
In a plane, the octet truss\cite{richard1961synergetic} is such a geometry, but in a line, the
Boerdijk--Coxeter helix is a regular structure.

@@

\begin{proof}
  Proof by our computer program that does this using Equation \eqref{eq:general}.
  See video on website\cite{readtetrahelix}.

\begin{figure}[H] %float with two figures
  \centering
     \includegraphics[width=0.4\textwidth]{figures/GlussBotBC.jpg}
     \caption{Glussbot in relaxed, or BC helix configuration}
     \includegraphics[width=0.4\textwidth]{figures/GlussBotEquitetrabeamCropped.jpg}
     \caption{The Equitetrabream: Fully Untwisted Glussbot}
\end{figure}
